Towards Practical Probabilistic Programming

### 2018-10-31 abstract/paper.tex
finished and squeezed

We believe that the probabilistic programming approach can be
made practical. However, this requires a shift from design of
languages best suited for probabilistic inference to design of
tools and libraries which make probabilistic programming in
modern general-purpose languages accessible and convenient.
Here, we first discuss the challenges in practical use of
existing probabilistic programming systems, and then suggest 
guidelines to implementation of a practical probabilistic
programming facility. We illustrate the guidelines on an
early-stage implementation of probabilistic programming facility
for the Go programming language. 
\end{abstract}

@@

Incorporating a probabilistic program, or rather a probabilistic
procedure, within a larger code body appears to be rather
straightforward: one implements the model in the probabilistic
programming language, fetches and preprocesses the data in the
`regular' host programming language, passes the data and the
model to an inference algorithm, and post-processes the results
in the `regular' programming language again to make algorithmic
decisions based on inference outcomes. The reality is more
complicated though, and the practitioner may face obstacles on
each step of the above routine.

\paragraph{Implementing the model} Probabilistic programming is
supposed to let the programmer use the expressive power of a
general programming language to specify probabilistic models.
However, the probabilistic language is often either a small
language lacking advanced control and data manipulation
capabilities (Anglican~\cite{TMY+16}, Stan~\cite{Stan17},
WebPPL~\cite{GS15}),
or an API to the representation of the probabilistic model
(Edward~\cite{THS+17}, Figaro~\cite{P09}). When the probabilistic
language is the same or as rich as the host language, it's use
is still constrained by available data structures
(Pyro~\cite{Pyro18}), or, the host language is itself uncommon as a
general-purpose language (Turing~\cite{GXG18}). Often, the
probabilistic programming language is interpreted, making
algorithm execution of the model prohibitively slow.

@@

\paragraph{Simulation-inference dichotomy} Many probabilistic
models follow a design pattern of simulation-inference: a
significant part of the model is a simulator, running an
algorithm with fixed parameter; the optimal parameters, or their
distribution, are to be inferred. The simulator may be quite
complicated, however in addition to the apparent difficulty of
implementing the simulator in a less expressive language,  the
simulator has to be implemented twice: first in the
probabilistic language for the inference, and then in the host
environment to make use of the inferred parameters. This not
just involves extra effort but is also prone to introducing
programming mistakes, when the two versions of the simulator
behave differently for the same parameter values. One possible
solution is to treat the simulator as a black box~\cite{LBW17},
however better inference algorithms can be used if the structure
of the simulator is revealed to the probabilistic programming
facility, e.g. through the use of automatic differentiation.

@@

Based on the experience of implementing  probabilistic
programming languages, as well as developing and deploying solutions
using different probabilistic environments, we suggest
the following guidelines to implementation of a probabilistic
programming facility. We believe that these guidelines, if
followed, will lead to eventual wide adoption of probabilistic
programming in software development practice.

@@

\section{Case Study: Probabilistic Programming Facility for
Go}

Following the guidelines, we have implemented a probabilistic
programming facility for the Go programming language,
\texttt{infergo} (\url{http://infergo.ml/}). At the time of
writing, the implementation is still at an early stage. However,
this facility is already used in production environment to infer
the website visitor's value using a hierarchical mixed effects
model.

>>>>

We believe that the probabilistic programming approach can be
made practical. However, this requires a shift from design of
languages best suited for probabilistic inference to design of
tools and libraries which make probabilistic programming in
modern general-purpose languages accessible and convenient.
Here, we first discuss the challenges in practical use of
existing probabilistic programming systems, and then suggest 
guidelines to implementation of a practical probabilistic
programming facility. We illustrate the guidelines on an
early-stage implementation of such facility for the Go
programming language. 
\end{abstract}

@@

Incorporating a probabilistic program, or rather a probabilistic
procedure, within a larger code body appears to be rather
straightforward: one implements the model in the probabilistic
language, fetches and preprocesses the data in the
`regular' host programming language, passes the data and the
model to an inference algorithm, and post-processes the results
in the `regular' programming language again to make algorithmic
decisions based on inference outcomes. The reality is more
complicated though, and the practitioner may face obstacles on
each step of the above routine.

\paragraph{Implementing the model} Probabilistic programming is
supposed to let the programmer use the expressive power of a
general programming language to specify probabilistic models.
However, the probabilistic language is often either a small
language lacking advanced control and data manipulation
capabilities (Anglican~\cite{TMY+16}, Stan~\cite{Stan17},
WebPPL~\cite{GS15}), or an API to the representation of the
probabilistic model (Edward~\cite{THS+17}, Figaro~\cite{P09}).
When the probabilistic language is the same or as rich as the
host language, its use is still constrained by available data
structures (Pyro~\cite{Pyro18}), or, the host language is 
uncommon as a general-purpose language (Turing~\cite{GXG18}).
Often, the probabilistic language is interpreted, making
execution of the model too slow.

@@

\paragraph{Simulation-inference dichotomy} Many probabilistic
models follow a design pattern of simulation-inference: a
significant part of the model is a simulator, running an
algorithm with fixed parameter; the optimal parameters, or their
distribution, are to be inferred. The simulator may be quite
complicated, however in addition to the apparent difficulty of
implementing the simulator in a less expressive language,  the
simulator has to be implemented twice: first in the
probabilistic language for inference, and then in the host
environment to make use of the inferred parameters. This not
just involves extra effort but is also prone to introducing
programming mistakes, when the two versions of the simulator
behave differently for the same parameter values. One possible
solution is to treat the simulator as a black box~\cite{LBW17},
however better inference algorithms can be used if the structure
of the simulator is revealed to the probabilistic programming
facility, e.g. through the use of automatic differentiation.

@@

Based on the experience of implementing  probabilistic
programming languages, as well as developing and deploying
solutions using different probabilistic environments, we suggest
guidelines to implementation of a probabilistic programming
facility. We believe that these guidelines, if followed, will
lead to eventual wide adoption of probabilistic programming in
software engineering.

@@

\section{Probabilistic Programming Facility for
Go}

Following the guidelines, we have implemented a probabilistic
programming facility for the Go programming language,
\texttt{infergo} (\url{http://infergo.ml/}). We have chosen Go
because Go is a relatively small programming language with
efficient implementation, which has recently become quite
popular for computation-intensive server-side programming. At
the time of writing, the implementation is still at an early
stage. However, this facility is already used in production
environment for inference of mission-critical algorithm
parameters.

A probabilistic model in \texttt{infergo} is an implementation
of the \texttt{Model} interface requiring a single method
\texttt{Observe} which accepts a vector (a Go \textit{slice}) of
floats, the parameters to infer, and returns a single float,
interpreted as unnormalized log-likelihood of the posterior
distribution. The implementation of the methods can be written
in virtually unrestricted Go and use any Go libraries.

For inference, \texttt{infergo} relies on automatic
differentiation. The source code of the model is
translated by a command-line tool provided by \texttt{infergo}
into an equivalent model with reverse-mode automatic
differentiation of the log-likelihood with respect 
to the parameters applied. The differentiation operates
on the built-in floating-point type and incurs only a small
computational overhead. However, even this overhead is avoided
when the model code is executed outside of inference algorithms:
both the original and the differentiated model are
simultaneously available to the rest of the program code, so
the methods can be called on the differentiated model for
inference, and on the original model for the most efficient
execution with the inferred parameters.

While we enjoy the ease and efficiency of
\texttt{infergo}-based inference in our production environment,
we view \texttt{infergo} as a reference implementation rather than
as the only solution, and encourage the community to equip other
general-purpose languages with similar inference facilities,
which, we believe, will make wide adoption of probabilistic
programming more likely.
be
